---
title: "3 Test Embedding"
author: "Rohan"
date: "2025-08-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dlookr)
library(purrr)
library(klaR)
library(caret)
library(nnet)
library(tidyr)
library(sampling)
library(dplyr)
library(ggplot2)
library(corrplot)
library(fastDummies)
library(factoextra)
library(cluster)
library(scales)
# Set seed for reproducibility
set.seed(123)
```

Defining global target variable, dataset_name and variables that will not go through encoding for the sake of reducing dimentionality
```{r}

# 1) Path to your CSV file:
dataset_path    <- "../Datasets/Merged_Dataset.csv"

# 2) Name of the column you want to predict (the “target”):
target_variable <- "Alzheimer.s.Diagnosis"

# 3) Any columns to drop before analysis (e.g. IDs, metadata):
exclude_variable <- character(0)

# If you don't need to drop anything, set:
# exclude_variable <- character(0) or else mention in "" and if more than one then c("","",.....)

# 4) Define your output folder relative to current wd
outdir <- "../All Reports/3. PCA and Clustering Embedding"
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
```


Creating user defined functions to Load Data

```{r}
# Function to load a CSV file into a data.frame
load_data <- function(path) {
  read.csv(
    path,
    stringsAsFactors = TRUE,    # keep character columns as characters
    na.strings      = c("", "NA", " ")  # treat these as missing
  )
}

# Load and inspect using the configured path
df <- load_data(dataset_path)
```


```{r}
# --- Your specs (must match the ones used in encoding) ---
binary1 <- c(
  "FamilyHistoryAlzheimers", "CardiovascularDisease", "Diabetes",
  "HeadInjury", "Hypertension", "MemoryComplaints",
  "BehavioralProblems", "Confusion", "Disorientation", "PersonalityChanges",
  "DifficultyCompletingTasks", "Forgetfulness", "Genetic.Risk.Factor..APOE.ε4.allele."
)

multi1 <- list(
  Gender         = list(levels = c(0,1),     labels = c("Male","Female")),
  EducationLevel = list(levels = c(0,1,2,3), labels = c("None","High School","Bachelor's","Higher")),
  Physical.Activity.CAT= list(levels = c(0,1,2),     labels = c("Low","Medium","High")),
  Smoking.Status= list(levels = c(0,1,2),     labels = c("Never","Former","Current")),
  Alcohol.Consumption.CAT= list(levels = c(0,1,2),     labels = c("Never", "Occasionally", "Regularly")),
  Cholesterol.Level= list(levels = c(0,1),     labels = c("Normal", "High")),
  Depression.Level= list(levels = c(0,1,2),     labels = c("Low", "Medium", "High")),
  Sleep.Quality.CAT= list(levels = c(0,1,2),     labels = c("Poor", "Average", "Good")),
  Diet.Quality.CAT= list(levels = c(0,1,2),     labels = c("Unhealthy", "Average", "Healthy")),
  Air.Pollution.Exposure= list(levels = c(0,1,2),     labels = c("Low", "Medium", "High")),
  Social.Engagement.Level= list(levels = c(0,1,2),     labels = c("Low", "Medium", "High")),
  Income.Level=  list(levels = c(0,1,2),     labels = c("Low", "Medium", "High")),
  Stress.Levels=  list(levels = c(0,1,2),     labels = c("Low", "Medium", "High"))
)

# --- Reverse function ---
revert_df <- function(df, binary_vars, multi_specs) {
  # Warn for missing binaries
  missing_bin <- setdiff(binary_vars, names(df))
  if (length(missing_bin)) {
    warning("These binary vars were not found and were skipped: ",
            paste(missing_bin, collapse = ", "), call. = FALSE)
  }

  # Revert binaries "No/Yes" -> 0/1 (skip if not present)
  df <- df %>%
    mutate(across(any_of(binary_vars),
                  ~ if (is.factor(.x) || is.character(.x)) {
                      c("No" = 0L, "Yes" = 1L)[as.character(.x)]
                    } else {
                      .x
                    }))

  # Warn for missing multi-levels
  missing_multi <- setdiff(names(multi_specs), names(df))
  if (length(missing_multi)) {
    warning("These multi-level vars were not found and were skipped: ",
            paste(missing_multi, collapse = ", "), call. = FALSE)
  }

  # Revert multi-levels using direct mapping
  for (var in intersect(names(multi_specs), names(df))) {
    specs <- multi_specs[[var]]
    mapping <- setNames(specs$levels, specs$labels)  # label -> numeric code
    df[[var]] <- if (is.factor(df[[var]]) || is.character(df[[var]])) {
      unname(mapping[as.character(df[[var]])])
    } else {
      df[[var]]
    }
  }

  df
}

# --- Usage ---
df <- revert_df(df, binary1, multi1)
```


```{r}
# One-hot encode all factor columns except the target and any exclusions, and convert dummies to factors
one_hot_encode_all_levels <- function(df, target_col, exclude_cols = exclude_variable) {
  # 1) Identify which columns to encode (all factors except target & excluded columns)
  cols_to_encode <- setdiff(
    names(df)[sapply(df, is.factor)],
    c(target_col, exclude_cols)
  )
  
  # 2) Use fastDummies to create dummy variables (keep all levels)
  df <- fastDummies::dummy_cols(
    df,
    select_columns          = cols_to_encode,
    remove_first_dummy      = FALSE,   # Keep all levels
    remove_selected_columns = TRUE     # Drop original factor columns
  )
  
  # 3) Convert the new dummy columns (0/1) to factors
  dummy_cols_created <- grep(paste0("^", paste(cols_to_encode, collapse = "|"), "_"), 
                             names(df), value = TRUE)
  df[dummy_cols_created] <- lapply(df[dummy_cols_created], factor)
  
  return(df)
}

# Apply one-hot encoding to your cleaned data
df_encoded <- one_hot_encode_all_levels(
  df,
  target_col  = target_variable,
  exclude_cols = exclude_variable
)

```

Working on a stratified sampling here as the dataset might be too large to run PCA on less powerful devices

```{r}
# Stratified sampling: creates a balanced sample by class (target variable)
stratified_sample_by_target <- function(df,
                                        target_col,
                                        sizes,
                                        remove_id = TRUE,
                                        id_col    = "ID") {
  # 1) Ensure the target column is a factor
  if (!is.factor(df[[target_col]])) {
    df[[target_col]] <- as.factor(df[[target_col]])
  }
  
  # 2) Get class labels and number of classes
  classes   <- levels(df[[target_col]])
  n_classes <- length(classes)
  
  # 3) Prepare the size vector for strata sampling
  if (length(sizes) == 1) {
    size_vec <- rep(sizes, n_classes)
  } else if (length(sizes) == n_classes) {
    size_vec <- sizes
  } else {
    stop("`sizes` must be either length 1 or equal to the number of classes (", 
         n_classes, ").")
  }
  
  # 4) Perform stratified sampling without replacement
  stratified_index <- sampling::strata(
    df,
    stratanames = target_col,
    size        = size_vec,
    method      = "srswor"
  )$ID_unit
  
  sampled_df <- df[stratified_index, ]
  
  # 5) Optionally drop the ID column
  if (remove_id && id_col %in% names(sampled_df)) {
    sampled_df <- sampled_df %>% dplyr::select(-all_of(id_col))
  }
  
  return(sampled_df)
}

# Example: Get a balanced sample (e.g., 3,000 per class)
df_ready <- stratified_sample_by_target(
  df         = df_encoded,
  target_col = target_variable,
  sizes      = 3260
)

# Check the resulting class proportions (should be balanced)
prop.table(table(df_ready[[target_variable]]))

```

```{r}
# Function to save a dlookr diagnose report with an “initial” prefix
save_diagnose_report <- function(df, name, prefix = "Initial") {
  report    <- df %>% diagnose()
  fname     <- paste0(prefix, "_", name, "_diagnose_report.csv")
  file_path <- file.path(outdir, fname)
  write.csv(report, file_path, row.names = FALSE)
  message("Saved report to ", file_path)
}

# Function to save summary report
save_summary_csv <- function(df, name, prefix = "Initial") {
  
  # 1) numeric summaries: min, Q1, median, mean, Q3, max
  num_vars  <- names(df)[sapply(df, is.numeric)]
  num_stats <- df %>%
    summarise(across(all_of(num_vars),
                     list(
                       min    = ~ min(.x,   na.rm = TRUE),
                       Q1     = ~ quantile(.x, .25, na.rm = TRUE),
                       median = ~ median(.x, na.rm = TRUE),
                       mean   = ~ mean(.x,   na.rm = TRUE),
                       Q3     = ~ quantile(.x, .75, na.rm = TRUE),
                       max    = ~ max(.x,   na.rm = TRUE)
                     ),
                     .names = "{col}_{fn}"
    )) %>%
    pivot_longer(
      cols      = everything(),
      names_to  = c("variable", "stat"),
      names_sep = "_"
    ) %>%
    pivot_wider(names_from = stat, values_from = value)
  
  # 2) factor summaries: number of levels + labels
  fac_vars  <- names(df)[sapply(df, is.factor)]
  fac_stats <- tibble(
    variable     = fac_vars,
    n_categories = map_int(fac_vars, ~ nlevels(df[[.x]])),
    categories   = map_chr(fac_vars, ~ paste(levels(df[[.x]]), collapse = ", "))
  )
  
  # 3) join into one table
  summary_tbl <- full_join(num_stats, fac_stats, by = "variable")
  
  # 4) write out
  fname <- paste0(prefix, "_", name, "_summary.csv")
  fpath <- file.path(outdir, fname)
  write.csv(summary_tbl, fpath, row.names = FALSE)
  message("Saved summary CSV to ", fpath)
}

# Usage

# Save dlookr diagnose report for the cleaned data
save_diagnose_report(df_ready, name = "dataset")

# Save summary CSV for the cleaned data
save_summary_csv(df_ready, name = "dataset")
```
```{r}
# Convert all binary factor columns to integers (0/1), leave others as is
convert_binary_factors_to_integers <- function(df, exclude_cols = NULL) {
  df[] <- lapply(names(df), function(colname) {
    col <- df[[colname]]
    if (is.factor(col) && nlevels(col) == 2 && !(colname %in% exclude_cols)) {
      # Convert factor to integer 0/1
      as.integer(as.numeric(col) - 1)
    } else {
      col
    }
  })
  names(df) <- names(df) # Preserve column names
  return(df)
}

# Example usage: exclude target variable
df_encoded <- convert_binary_factors_to_integers(df_ready, exclude_cols = target_variable)

```


```{r}
# Function: Standardise numeric variables (Z-score) except binary and target
standardise_numeric_except_binary <- function(df, target_col) {
  df[] <- lapply(names(df), function(col_name) {
    col <- df[[col_name]]
    
    # Leave non-numeric columns unchanged
    if (!is.numeric(col)) return(col)
    
    # Leave target column unchanged
    if (col_name == target_col) return(col)
    
    # Leave binary numeric variables unchanged (only 0 and 1 present)
    unique_vals <- unique(col[!is.na(col)])
    if (length(unique_vals) <= 2 && all(unique_vals %in% c(0, 1))) {
      return(col)
    }
    
    # Apply Z-score standardisation
    mu <- mean(col, na.rm = TRUE)
    sigma <- sd(col, na.rm = TRUE)
    if (sigma == 0 || is.na(sigma)) {
      return(col)  # avoid division by zero
    } else {
      return((col - mu) / sigma)
    }
  })
  
  # Return as data frame
  df <- as.data.frame(df)
  return(df)
}

# ==== Example usage ====
df_scaled <- standardise_numeric_except_binary(df_encoded, target_col = target_variable)

```

```{r}
str(df_scaled)
```


PCA
```{r}
# Run PCA on all numeric columns except the target and excluded columns
run_pca <- function(df, target_col, exclude_cols = exclude_variable) {
  # 1) Determine columns to exclude (target and any additional exclusions)
  cols_to_exclude <- c(target_col, exclude_cols)
  
  # 2) Subset only columns not in the exclude list
  num_df <- df[, !(names(df) %in% cols_to_exclude)]
  
  # 3) Keep only numeric columns
  num_df <- num_df[, sapply(num_df, is.numeric)]
  
  # 4) Scale (standardize) data and run PCA
  pca <- prcomp(num_df, center = TRUE, scale. = FALSE)
  
  # 5) Print explained variance for each principal component
  variance <- summary(pca)$importance[2, ]
  cat("Explained variance by component:\n")
  print(variance)
  
  return(pca)
}

# Example: Run PCA on your encoded dataframe
pca_result <- run_pca(
  df_scaled,
  target_col    = target_variable,
  exclude_cols  = exclude_variable
)
```

```{r}
# Plot a scree plot for a PCA result object
plot_pca_scree <- function(pca) {
  # 1) Calculate the proportion of variance explained by each PC
  var_explained <- pca$sdev^2 / sum(pca$sdev^2)
  
  # 2) Create a data frame for plotting
  scree_data <- data.frame(
    PC = seq_along(var_explained),
    Variance = var_explained
  )
  
  # 3) Plot the scree plot with ggplot2
  ggplot(scree_data, aes(x = PC, y = Variance)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "PCA Scree Plot",
      x     = "Principal Component",
      y     = "Proportion of Variance Explained"
    ) +
    theme_minimal()
}

# Call the function to display the scree plot
plot_pca_scree(pca_result)

```


```{r}
# Run PCA and plot cumulative explained variance
run_pca_analysis <- function(df, target_col = target_variable, exclude_cols = exclude_variable) {
  # 1) Select only numeric columns, then drop the target and any excluded columns
  num_df <- df %>%
    dplyr::select(where(is.numeric)) %>%
    dplyr::select(-dplyr::any_of(c(target_col, exclude_cols)))
  
  # 2) Run PCA with centering and scaling
  pca_result <- prcomp(num_df, center = TRUE, scale. = FALSE)
  
  # 3) Print summary of PCA (shows variance explained, etc.)
  print(summary(pca_result))
  
  # 4) Calculate and store the proportion of explained variance (PEV) for each PC
  pca_result_var <- pca_result$sdev^2
  pca_result_PEV <- pca_result_var / sum(pca_result_var)
  
  # 5) Plot cumulative explained variance (scree-style, but cumulative)
  opar <- par(no.readonly = TRUE)   # Save old graphics parameters
  plot(
    cumsum(pca_result_PEV),
    ylim = c(0, 1),
    xlab = 'Principal Component',
    ylab = 'Cumulative Proportion of Explained Variance',
    type = "b",
    col  = "orange",
    main = "Cumulative Explained Variance by PCs"
  )
  abline(h = 0.8, col = 'red', lty = 'dashed')  # 80% threshold reference
  par(opar)   # Restore old graphics parameters
  
  # 6) Return the PCA object for further analysis (scores, loadings, etc.)
  return(pca_result)
}

# Example usage:
pca_result <- run_pca_analysis(
  df_scaled,
  target_col   = target_variable,
  exclude_cols = exclude_variable
)

```

K-means Clustering

```{r}

#Run PCA on the stratified sample instead of the full encoded set
pca_result_ready <- run_pca(
  df         = df_scaled,           #stratified sample
  target_col = target_variable,
  exclude_cols = exclude_variable
)

#Decide how many PCs to keep (90% variance considered)
var_explained <- summary(pca_result_ready)$importance[2, ]
cum_var       <- cumsum(var_explained)
threshold     <- 0.90
n_pcs         <- which(cum_var >= threshold)[1]
cat(
  "Retaining", n_pcs, "PCs (",
  round(cum_var[n_pcs] * 100, 1), "% variance)\n"
)

#Considering those top‐PC scores
pca_scores_ready <- pca_result_ready$x[, 1:n_pcs, drop = FALSE]

#Elbow method on the entire stratified sample
fviz_nbclust(
  pca_scores_ready,
  FUN     = kmeans,
  method  = "wss",
  k.max   = 10
) +
  labs(title = "Elbow Method on Stratified Sample") +
  theme_classic()

#Silhouette analysis on the entire stratified sample
k_values        <- 2:9
kmeans_results  <- list()
sil_widths      <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  
  km <- kmeans(pca_scores_ready, centers = k)
  
  # compute silhouette
  sil <- silhouette(km$cluster, dist(pca_scores_ready))
  sil_widths[i] <- mean(sil[, "sil_width"])
  
  #plot
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ")"),
    border = NA,
    cex.names = 0.6
  )
  
  kmeans_results[[paste0("k", k)]] <- km
}

#Compare average silhouette widths to pick the best k
sil_df <- data.frame(
  k   = k_values,
  ASW = sil_widths
)
print(sil_df)

# Build silhouette summary
sil_df <- data.frame(
  k   = k_values,
  ASW = sil_widths
)
```

```{r}
# 1. Pick the best number of clusters (k) based on silhouette analysis
best_k <- sil_df$k[which.max(sil_df$ASW)]
cat("Best k by average silhouette:", best_k, "\n")

# 2. Re-run k-means clustering using the optimal k on the top principal components
km_best <- kmeans(pca_scores_ready, centers = best_k)

# 3. Prepare a data frame for plotting: use first two PCs and cluster assignments
plot_df <- data.frame(
  PC1     = pca_scores_ready[, 1],
  PC2     = pca_scores_ready[, 2],
  cluster = factor(km_best$cluster)
)

# 4. Scatter plot the clusters colored by their cluster assignment
ggplot(plot_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("K-means Clustering (k = ", best_k, ") on First Two PCs"),
    x     = "Principal Component 1",
    y     = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```

------K-means Clustering without PCA
```{r}
# 1. Prepare data: remove target/excluded columns and apply min-max scaling
orig_clust_df <- df_scaled %>%
  dplyr::select(-dplyr::any_of(c(target_variable, exclude_variable)))
orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

# 2. K-means silhouette analysis for k = 2 to 10
library(cluster)
sil_widths_km <- numeric()
kmeans_results <- list()
for (k in 2:10) {
  set.seed(123)  # for reproducibility
  km <- kmeans(orig_clust_minmax, centers = k)
  kmeans_results[[k]] <- km
  sil <- silhouette(km$cluster, dist(orig_clust_minmax))
  sil_widths_km[k] <- mean(sil[, "sil_width"])
  # Plot silhouette for each k
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ") - K-means (Original Features)"),
    border = NA,
    cex.names = 0.6
  )
}

# 3. Silhouette summary table for all k
sil_df_km <- data.frame(
  k   = 2:10,
  ASW = sil_widths_km[2:10]
)
print(sil_df_km)

# 4. Plot average silhouette width vs k
plot(sil_df_km$k, sil_df_km$ASW, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Width vs. k (K-means, Original Features)")

# 5. Select best k (highest silhouette width)
best_k_km <- sil_df_km$k[which.max(sil_df_km$ASW)]
cat("Best k by average silhouette width (k-means):", best_k_km, "\n")

# 6. Run final k-means on original data with best k
set.seed(123)
km_best <- kmeans(orig_clust_minmax, centers = best_k_km)

# 7. Plot clusters on the first two min-max scaled features (no PCA)
km_plot_df <- data.frame(
  Feature1 = orig_clust_minmax[, 1],
  Feature2 = orig_clust_minmax[, 2],
  cluster  = factor(km_best$cluster)
)

library(ggplot2)
ggplot(km_plot_df, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("K-means Clustering (k = ", best_k_km, ") on First Two Features"),
    x     = "Feature 1",
    y     = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```

TEST CLUSTERING WITHOUT PCA RESULT
```{r}
#── 1. Prepare data: drop target/excluded and Min–Max scale ────────────────────
orig_clust_df <- df_scaled %>%
  select(- any_of(c(target_variable, exclude_variable)))

orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

#── 2. Silhouette analysis for k = 2:10 ─────────────────────────────────────────
sil_widths_km <- numeric(10)
kmeans_results <- vector("list", 10)

for (k in 2:10) {
  set.seed(123)  
  km <- kmeans(orig_clust_minmax, centers = k)
  kmeans_results[[k]] <- km

  sil <- silhouette(km$cluster, dist(orig_clust_minmax))
  sil_widths_km[k] <- mean(sil[, "sil_width"])
  
  # optional: display each silhouette plot
  plot(sil,
       main = paste("Silhouette Plot (k =", k, ")"),
       border = NA, cex.names = 0.6)
}

#── 3. Summarize silhouette widths and pick best k ─────────────────────────────
sil_df_km <- data.frame(
  k   = 2:10,
  ASW = sil_widths_km[2:10]
)
print(sil_df_km)

best_k_km <- sil_df_km$k[which.max(sil_df_km$ASW)]
cat("Best k by average silhouette width:", best_k_km, "\n")

#── 4. Final k-means with best k ────────────────────────────────────────────────
set.seed(123)
km_best <- kmeans(orig_clust_minmax, centers = best_k_km)

#── 5. (Optional) Check cardinality of first two raw features ───────────────────
cat("Unique values in raw feature 1 & 2:\n")
print(sapply(orig_clust_df[, 1:2], function(x) length(unique(x))))

#── 6. Plot clustered data on first two features *with jitter* ─────────────────
km_plot_df <- data.frame(
  Feature1 = orig_clust_minmax[[1]],
  Feature2 = orig_clust_minmax[[2]],
  cluster  = factor(km_best$cluster)
)

ggplot(km_plot_df, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_jitter(width = 0.01, height = 0.01, alpha = 0.6, size = 1.5) +
  labs(
    title = paste0("K‑means Clustering (k=", best_k_km, ") on Features 1 & 2 (jittered)"),
    x     = "Feature 1 (min–max scaled)",
    y     = "Feature 2 (min–max scaled)",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

#── 7. (Optional) 2D PCA plot for comparison ────────────────────────────────────
pca_res <- prcomp(orig_clust_minmax, center = TRUE, scale. = FALSE)
pca_df <- data.frame(
  PC1     = pca_res$x[, 1],
  PC2     = pca_res$x[, 2],
  cluster = factor(km_best$cluster)
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("K‑means Clustering (k=", best_k_km, ") on PC1 vs PC2"),
    x     = "PC1",
    y     = "PC2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )
```

----- Clustering Hierarchical

```{r}
# Hierarchical clustering on the stratified PCA scores

# 1. Compute the Euclidean distance matrix on the PCA scores
hc_dist_full <- dist(pca_scores_ready, method = "euclidean")

# 2. Perform hierarchical clustering using Ward's method
hc_ward_full <- hclust(hc_dist_full, method = "ward.D2")

# 3. Plot the dendrogram (labels omitted for clarity)
plot(
  as.dendrogram(hc_ward_full),
  main   = "Hierarchical Clustering Dendrogram (Ward) on Full Sample",
  ylab   = "Height",
  labels = FALSE,
  hang   = -1
)

# 4. Highlight the best_k clusters on the dendrogram
rect.hclust(hc_ward_full, k = best_k, border = "steelblue")

# 5. Cut the tree into best_k clusters
hc_clusters_full <- cutree(hc_ward_full, k = best_k)

# 6. Prepare data for scatter plot: first two principal components, colored by cluster
hc_plot_df_full <- data.frame(
  PC1     = pca_scores_ready[, 1],
  PC2     = pca_scores_ready[, 2],
  cluster = factor(hc_clusters_full)
)

# 7. Scatter plot of clusters on first two PCs
ggplot(hc_plot_df_full, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k = ", best_k, ") on Full Stratified PCs"),
    x     = "Principal Component 1",
    y     = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```

----- Clustering Hierarchical without PCA
```{r}
# 1. Prepare data: remove target/excluded columns and apply min-max scaling
orig_clust_df <- df_scaled %>%
  dplyr::select(-dplyr::any_of(c(target_variable, exclude_variable)))
orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

# 2. Compute distance matrix and hierarchical clustering
hc_dist_orig <- dist(orig_clust_minmax, method = "euclidean")
hc_ward_orig <- hclust(hc_dist_orig, method = "ward.D2")

# 3. Silhouette analysis for k = 2 to 10
library(cluster)
sil_widths_hc <- numeric()
for (k in 2:10) {
  clusters <- cutree(hc_ward_orig, k)
  sil <- silhouette(clusters, hc_dist_orig)
  sil_widths_hc[k] <- mean(sil[, "sil_width"])
  # Plot silhouette for each k
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ") - Hierarchical (Original Features)"),
    border = NA,
    cex.names = 0.6
  )
}

# 4. Plot average silhouette width vs. k
plot(2:10, sil_widths_hc[2:10], type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Width vs. k (Hierarchical, Original Features)")

# 5. Select best k (highest silhouette width)
best_k_hc <- which.max(sil_widths_hc)
cat("Best k by average silhouette width (hierarchical):", best_k_hc, "\n")

# 6. Dendrogram with rectangles for chosen clusters
plot(
  as.dendrogram(hc_ward_orig),
  main   = paste0("Hierarchical Clustering Dendrogram (k = ", best_k_hc, ")"),
  ylab   = "Height",
  labels = FALSE,
  hang   = -1
)
rect.hclust(hc_ward_orig, k = best_k_hc, border = "steelblue")

# 7. Assign clusters for the best k
hc_clusters_orig <- cutree(hc_ward_orig, k = best_k_hc)

# 8. Plot clusters on the first two min-max scaled features (no PCA)
hc_plot_df_orig <- data.frame(
  Feature1 = orig_clust_minmax[, 1],
  Feature2 = orig_clust_minmax[, 2],
  cluster  = factor(hc_clusters_orig)
)

library(ggplot2)
ggplot(hc_plot_df_orig, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k = ", best_k_hc, ") on First Two Features"),
    x     = "Feature 1",
    y     = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```

TEST CLUSTERING WITHOUT PCA RESULT
```{r}
#── 1. Prepare data: drop target/excluded and Min–Max scale ────────────────────
orig_clust_df <- df_scaled %>%
  select(- any_of(c(target_variable, exclude_variable)))

orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

#── 2. Compute distance matrix and hierarchical clustering (Ward’s method) ────
hc_dist_orig  <- dist(orig_clust_minmax, method = "euclidean")
hc_ward_orig  <- hclust(hc_dist_orig, method = "ward.D2")

#── 3. Silhouette analysis for k = 2:10 ────────────────────────────────────────
sil_widths_hc <- numeric(10)

for (k in 2:10) {
  clusters <- cutree(hc_ward_orig, k)
  sil      <- silhouette(clusters, hc_dist_orig)
  sil_widths_hc[k] <- mean(sil[, "sil_width"])
  
  # optional: display each silhouette plot
  plot(sil,
       main = paste("Silhouette Plot (k =", k, ") - Hierarchical"),
       border = NA, cex.names = 0.6)
}

#── 4. Summarize silhouette widths and pick best k ─────────────────────────────
sil_df_hc <- data.frame(
  k   = 2:10,
  ASW = sil_widths_hc[2:10]
)
print(sil_df_hc)

best_k_hc <- sil_df_hc$k[which.max(sil_df_hc$ASW)]
cat("Best k by average silhouette width (hierarchical):", best_k_hc, "\n")

#── 5. Plot Avg. Silhouette Width vs k ────────────────────────────────────────
plot(sil_df_hc$k, sil_df_hc$ASW, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Width vs. k (Hierarchical)")

#── 6. Dendrogram with rectangles for chosen clusters ─────────────────────────
plot(as.dendrogram(hc_ward_orig),
     main   = paste0("Hierarchical Clustering Dendrogram (k = ", best_k_hc, ")"),
     ylab   = "Height",
     labels = FALSE,
     hang   = -1)
rect.hclust(hc_ward_orig, k = best_k_hc, border = "steelblue")

#── 7. Assign clusters for the best k ─────────────────────────────────────────
hc_clusters_orig <- cutree(hc_ward_orig, k = best_k_hc)

#── 8. (Optional) Check cardinality of first two raw features ──────────────────
cat("Unique values in raw feature 1 & 2:\n")
print(sapply(orig_clust_df[, 1:2], function(x) length(unique(x))))

#── 9. Plot clustered data on first two features *with jitter* ────────────────
hc_plot_df_orig <- data.frame(
  Feature1 = orig_clust_minmax[[1]],
  Feature2 = orig_clust_minmax[[2]],
  cluster  = factor(hc_clusters_orig)
)

ggplot(hc_plot_df_orig, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_jitter(width = 0.01, height = 0.01, alpha = 0.6, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k=", best_k_hc, 
                   ") on Features 1 & 2 (jittered)"),
    x     = "Feature 1 (min–max scaled)",
    y     = "Feature 2 (min–max scaled)",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

#── 10. (Optional) 2D PCA plot for comparison ─────────────────────────────────
pca_res <- prcomp(orig_clust_minmax, center = TRUE, scale. = FALSE)
pca_df  <- data.frame(
  PC1     = pca_res$x[, 1],
  PC2     = pca_res$x[, 2],
  cluster = factor(hc_clusters_orig)
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k=", best_k_hc, 
                   ") on PC1 vs PC2"),
    x     = "PC1",
    y     = "PC2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )
```

```{r}
# Packages (install if needed)
# install.packages(c("caret","randomForest","pROC"))
library(caret)
library(randomForest)
library(pROC)
library(dplyr)

set.seed(123)

# 0) Copy & make a clean binary target called 'Alz' with levels No/Yes
df_rf <- df_scaled

# robust target -> Alz mapping (handles 0/1 or Yes/No)
tcol <- target_variable
if (is.numeric(df_rf[[tcol]])) {
  df_rf$Alz <- factor(ifelse(df_rf[[tcol]] == 1, "Yes", "No"),
                      levels = c("No","Yes"))
} else {
  v <- as.character(df_rf[[tcol]])
  yes_vals <- c("1","yes","Yes","Y","Positive","AD","Alzheimer")
  df_rf$Alz <- factor(ifelse(v %in% yes_vals, "Yes", "No"),
                      levels = c("No","Yes"))
}

# 1) Drop the original target column from predictors
df_rf <- df_rf %>% select(-all_of(tcol))

# 2) Remove near-zero-variance predictors (important after encoding)
nzv <- nearZeroVar(df_rf)
if (length(nzv) > 0) df_rf <- df_rf[, -nzv]

# 3) Train/test split stratified on Alz
set.seed(123)
idx <- createDataPartition(df_rf$Alz, p = 0.7, list = FALSE)
train_pp <- df_rf[idx, ]
test_pp  <- df_rf[-idx, ]

# 4) TrainControl for ROC
ctrl <- trainControl(
  method          = "repeatedcv",
  number          = 5,
  repeats         = 2,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# 5) Random Forest (single mtry like your example). Adjust mtry if needed.
p <- ncol(train_pp) - 1  # predictors count
rf_mod <- train(
  Alz ~ .,
  data      = train_pp,
  method    = "rf",
  metric    = "ROC",
  trControl = ctrl,
  tuneGrid  = data.frame(mtry = floor(sqrt(p))),  # or set a specific value, e.g. 5
  ntree     = 500
)

print(rf_mod)

# 6) Evaluate on hold-out
rf_probs <- predict(rf_mod, test_pp, type = "prob")[, "Yes"]
rf_preds <- ifelse(rf_probs > 0.5, "Yes", "No")

rf_cm <- confusionMatrix(
  factor(rf_preds, levels = c("No","Yes")),
  test_pp$Alz,
  positive = "Yes"
)
print(rf_cm)

# Accuracy
rf_acc <- rf_cm$overall["Accuracy"]
cat("RF Accuracy:", round(rf_acc, 4), "\n")

# AUC
rf_roc <- roc(
  response  = test_pp$Alz,
  predictor = rf_probs,
  levels    = c("No","Yes"),
  direction = "<"
)
cat("RF AUC:", round(auc(rf_roc), 4), "\n")
plot(rf_roc, main = "ROC Curve — Random Forest")

# 7) (Optional) Variable importance
vip <- varImp(rf_mod, scale = TRUE)
print(vip)
plot(vip, top = 20, main = "Top 20 Important Predictors (RF)")

```

```{r}
# install.packages(c("caret","xgboost","pROC")) # if needed
library(caret)
library(pROC)
library(dplyr)

set.seed(123)

# 0) Prepare data and Alz target (No/Yes)
df_xgb <- df_scaled
tcol <- target_variable
if (is.numeric(df_xgb[[tcol]])) {
  df_xgb$Alz <- factor(ifelse(df_xgb[[tcol]] == 1, "Yes", "No"),
                       levels = c("No","Yes"))
} else {
  v <- as.character(df_xgb[[tcol]])
  yes_vals <- c("1","yes","Yes","Y","Positive","AD","Alzheimer")
  df_xgb$Alz <- factor(ifelse(v %in% yes_vals, "Yes", "No"),
                       levels = c("No","Yes"))
}
df_xgb <- df_xgb %>% select(-all_of(tcol))  # drop original target

# 1) (Optional) remove near-zero-variance predictors to speed up
nzv <- nearZeroVar(df_xgb)
if (length(nzv) > 0) df_xgb <- df_xgb[, -nzv]

# 2) Train/test split
set.seed(123)
idx <- createDataPartition(df_xgb$Alz, p = 0.7, list = FALSE)
train_pp <- df_xgb[idx, ]
test_pp  <- df_xgb[-idx, ]

# 3) Lightweight trainControl (3-fold CV, ROC metric)
ctrl <- trainControl(
  method          = "cv",
  number          = 3,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  verboseIter     = FALSE
)

# --- XGBoost (single grid row, fast) ---
xgb_mod <- train(
  Alz ~ .,
  data      = train_pp,
  method    = "xgbTree",
  metric    = "ROC",
  trControl = ctrl,
  tuneGrid  = data.frame(
    nrounds          = 100,
    max_depth        = 6,
    eta              = 0.3,
    gamma            = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample        = 0.8
  )
)

print(xgb_mod)

# 4) Evaluate on the hold-out
xgb_probs <- predict(xgb_mod, test_pp, type = "prob")[, "Yes"]
xgb_preds <- ifelse(xgb_probs > 0.5, "Yes", "No")

xgb_cm  <- confusionMatrix(
  factor(xgb_preds, levels = c("No","Yes")),
  test_pp$Alz,
  positive = "Yes"
)
print(xgb_cm)

# Accuracy
xgb_acc <- xgb_cm$overall["Accuracy"]
cat("XGB Accuracy:", round(xgb_acc, 4), "\n")

# AUC + ROC
xgb_roc <- roc(
  response  = test_pp$Alz,
  predictor = xgb_probs,
  levels    = c("No","Yes")
)
cat("XGB AUC:", round(auc(xgb_roc), 4), "\n")
plot(xgb_roc, main = "ROC Curve for XGBoost (single-grid)")

```

```{r}
# install.packages(c("caret","gbm","pROC")) # if needed
library(caret)
library(gbm)
library(pROC)

# --- GBM with caret's default tuning ---
gbm_mod <- train(
  Alz ~ .,                      # or: Alz ~ . - Alzheimer.s.Diagnosis
  data      = train_pp,
  method    = "gbm",
  metric    = "ROC",
  trControl = ctrl,              # e.g., 3-fold CV with classProbs + twoClassSummary
  verbose   = FALSE              # suppress gbm training output
  # no tuneGrid → caret explores its default grid
)

print(gbm_mod)

# --- Evaluate on hold-out ---
gbm_probs <- predict(gbm_mod, test_pp, type = "prob")[, "Yes"]
gbm_preds <- ifelse(gbm_probs > 0.5, "Yes", "No")

gbm_cm <- confusionMatrix(
  factor(gbm_preds, levels = c("No","Yes")),
  test_pp$Alz,
  positive = "Yes"
)
print(gbm_cm)
cat("GBM Accuracy:", round(gbm_cm$overall["Accuracy"], 4), "\n")

# AUC & ROC
gbm_roc <- roc(
  response  = test_pp$Alz,
  predictor = gbm_probs,
  levels    = c("No","Yes"),
  direction = "<"
)
cat("GBM AUC:", round(auc(gbm_roc), 4), "\n")
plot(gbm_roc, main = "ROC Curve for GBM")

```

```{r}
# install.packages(c("caret","C50","pROC")) # if needed
library(caret)
library(C50)
library(pROC)

# --- C5.0 with caret's default tuning ---
c50_mod <- train(
  Alz ~ .,                      # or: Alz ~ . - Alzheimer.s.Diagnosis
  data      = train_pp,
  method    = "C5.0",
  metric    = "ROC",
  trControl = ctrl               # e.g., 3-fold CV with classProbs + twoClassSummary
  # no tuneGrid here → caret will explore the default grid of trials/model/winnow
)

print(c50_mod)

# --- Evaluate on hold-out ---
c50_probs <- predict(c50_mod, test_pp, type = "prob")[, "Yes"]
c50_preds <- ifelse(c50_probs > 0.5, "Yes", "No")

c50_cm <- confusionMatrix(
  factor(c50_preds, levels = c("No","Yes")),
  test_pp$Alz,
  positive = "Yes"
)
print(c50_cm)
cat("C5.0 Accuracy:", round(c50_cm$overall["Accuracy"], 4), "\n")

# AUC & ROC
c50_roc <- roc(
  response  = test_pp$Alz,
  predictor = c50_probs,
  levels    = c("No","Yes"),
  direction = "<"
)
cat("C5.0 AUC:", round(auc(c50_roc), 4), "\n")
plot(c50_roc, main = "ROC Curve for C5.0")

```
