---
title: "Alzheimers R Code"
author: "Rohan"
date: "2025-07-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(sampling)
library(dplyr)
library(ggplot2)
library(corrplot)
library(fastDummies)
library(factoextra)
library(cluster)
library(scales)
library(dlookr)
set.seed(123)
```

Defining global target variable, dataset_name and variables that will not go through encoding for the sake of reducing dimentionality
```{r}


# 1) Path to your CSV file:
dataset_path    <- "../Datasets/Merged_Dataset.csv"

# 2) Name of the column you want to predict (the “target”):
target_variable <- "Alzheimer.s.Diagnosis"

# 3) Any columns to drop before analysis (e.g. IDs, metadata):
exclude_variable <- character(0)

# If you don't need to drop anything, set:
# exclude_variable <- character(0) or else mention in "" and if more than one then c("","",.....)

```


Creating user defined functions to Load and Inspect Data

```{r}

# Function to load a CSV file into a data.frame
load_data <- function(path) {
  read.csv(
    path,
    stringsAsFactors = FALSE,    # keep character columns as characters
    na.strings      = c("", "NA", " ")  # treat these as missing
  )
}

# Function to print basic info about a data.frame
inspect_structure <- function(df) {
  cat("Dimensions:", dim(df), "\n")                     # rows x columns
  cat("Column names:\n"); print(names(df))              # list of columns
  cat("Summary statistics:\n"); print(summary(df))      # min, max, median, etc.
  cat("Missing values per column:\n"); 
  print(colSums(is.na(df)))                             # count of NAs
  cat("Data types:\n"); print(sapply(df, class))        # class of each column
}

# Load and inspect using the configured path
df <- load_data(dataset_path)   # load your CSV into 'df'

# Drop any duplicate columns ending with ".-"
df <- df[, !grepl("\\.-$", names(df))]

inspect_structure(df)           # display its structure and summary

```


```{r}

# Compute mode (works for vectors)
get_mode <- function(v) {
  uniqv <- unique(na.omit(v))
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Clean missing values & coerce numeric-like text to numeric using dplyr
clean_missing_values <- function(df) {
  df <- df %>%
    mutate(across(
      where(is.character),
      ~{
        clean <- gsub("[^0-9\\.-]", "", .)           # remove non-numeric characters
        clean[clean == ""] <- NA                    # turn empty strings into NA
        num <- suppressWarnings(as.numeric(clean))   # try converting to numeric
        if (any(!is.na(num))) num else .           # use numeric if possible, otherwise original
      }
    )) %>%
    mutate(across(
      where(is.numeric),
      ~replace(., is.na(.), median(., na.rm = TRUE))  # replacing NA values for numerical columns with median
    )) %>%
    mutate(across(
      where(~is.factor(.) || is.character(.)),  # replacing NA values for categorical columns with mode
      ~replace(., is.na(.), get_mode(.))
    ))
  return(df)
}

# Replace outliers with median for all numeric columns using dplyr
replace_outliers_with_median <- function(df) {
  num_cols <- names(df)[sapply(df, is.numeric)]
  for (col in num_cols) {
    out_vals <- boxplot(df[[col]], plot = FALSE)$out
    df[[col]][df[[col]] %in% out_vals] <- NA
    df[[col]][is.na(df[[col]])] <- median(df[[col]], na.rm = TRUE)
  }
  return(df)
}

# Convert all remaining character columns to factors using dplyr
convert_char_to_factor <- function(df) {
  df <- df %>%
    mutate(across(where(is.character), as.factor))
  return(df)
}

# --- EXECUTE CLEANING PIPELINE ---
df <- df %>%
  clean_missing_values() %>%
  replace_outliers_with_median() %>%
  convert_char_to_factor()

# 4. Drop any columns listed in exclude_variable
if (length(exclude_variable) > 0) {
  df <- df %>% select(-any_of(exclude_variable))
}

# Display the structure of the cleaned dataframe
str(df)

```


```{r}
#change visual of boxplot
# Boxplots of each numeric variable grouped by the (now factor) target variable
plot_boxplots_by_target <- function(df, target_col) {
  # Make a local copy so we don't modify the original dataframe
  plot_df <- df
  
  # 1) Ensure the target is a factor (for grouping)
  plot_df[[target_col]] <- as.factor(plot_df[[target_col]])
  
  # 2) Identify all numeric columns (excluding the target variable itself)
  num_vars <- setdiff(
    names(plot_df)[sapply(plot_df, is.numeric)],
    target_col
  )
  
  # 3) Loop over each numeric variable and print its boxplot by target group
  for (var in num_vars) {
    p <- ggplot(plot_df, aes(x = .data[[target_col]], y = .data[[var]])) +
      geom_boxplot(fill = "lightblue", outlier.alpha = 0.3) +
      theme_minimal() +
      labs(
        title = paste("Boxplot of", var, "by", target_col),
        x     = target_col,
        y     = var
      )
    print(p)
  }
}

# Call the function with your pre-defined target_variable
plot_boxplots_by_target(df, target_variable)

```

```{r}
#visual change, add density curve
# Plot histograms for each numeric column in the dataframe
plot_numeric_distributions <- function(df) {
  # 1) Identify all numeric columns in the dataframe
  num_vars <- names(df)[sapply(df, is.numeric)]
  
  # 2) Loop through each numeric variable and plot its histogram
  for (var in num_vars) {
    p <- ggplot(df, aes(x = .data[[var]])) +
      geom_histogram(bins = 30, fill = "steelblue", color = "white") +
      theme_minimal() +
      labs(
        title = paste("Histogram of", var),
        x     = var,
        y     = "Count"
      )
    print(p)  # Only display the plot, do not save
  }
}

# Call the function to plot all numeric distributions
plot_numeric_distributions(df)


```

```{r}
# 1. Specify columns to exclude from correlation analysis
exclude_cols <- c(target_variable, "DiagnosisBinary", "ID")  # Add more column names if needed

# 2. Remove excluded columns
df_corr <- df %>% dplyr::select(-dplyr::any_of(exclude_cols))

# 3. Coerce all numeric and factor columns to numeric
df_num <- df_corr %>%
  mutate(across(
    where(~ is.numeric(.) || is.factor(.)),
    ~as.numeric(as.character(.))
  ))

# 4. Keep only columns with more than one unique (non-NA) value
var_cols <- sapply(df_num, function(x) is.numeric(x) && length(unique(na.omit(x))) > 1)
df_num   <- df_num[, var_cols, drop = FALSE]

# 5. Check if enough variables exist for correlation
if (ncol(df_num) < 2) {
  cat("Not enough numeric columns with variation for correlation analysis.\n")
} else {
  # 6. Compute the correlation matrix
  cm <- cor(df_num, use = "pairwise.complete.obs")
  
  # 7. Only look at unique pairs (remove lower triangle and diagonal)
  cm[lower.tri(cm, diag = TRUE)] <- NA
  
  # 8. Identify variable pairs with abs(correlation) > 0.7
  corr_pairs <- which(abs(cm) > 0.7, arr.ind = TRUE)
  
  # 9. Print variable pairs with high correlation
  if (nrow(corr_pairs) == 0) {
    cat("No variable pairs with |correlation| > 0.7\n")
  } else {
    result <- data.frame(
      Var1 = rownames(cm)[corr_pairs[,1]],
      Var2 = colnames(cm)[corr_pairs[,2]],
      Correlation = cm[corr_pairs]
    )
    print(result)
  }
}

```




```{r}
# 1. List columns to exclude (target, binary target, IDs, or others)
exclude_cols <- c(target_variable, "DiagnosisBinary", "ID")  # Add others if needed

# 2. Remove excluded columns before correlation
df_corr <- df %>% dplyr::select(-dplyr::any_of(exclude_cols))

# 3. Convert all numeric and factor columns to numeric (so factors are coded as numbers)
df_num <- df_corr %>%
  mutate(across(
    where(~ is.numeric(.) || is.factor(.)),
    ~as.numeric(as.character(.))
  ))

# 4. Keep only columns that are numeric AND have more than one unique non-NA value
var_cols <- sapply(df_num, function(x) is.numeric(x) && length(unique(na.omit(x))) > 1)
df_num   <- df_num[, var_cols, drop = FALSE]

# 5. Compute the correlation matrix on these columns
cm <- cor(df_num, use = "pairwise.complete.obs")

# 6. Print the correlation matrix
print(cm)

```


Correlation matrix for numeric variables
```{r}
plot_correlation_matrix <- function(df, exclude_cols = c(target_variable, "DiagnosisBinary", "ID")) {
  # 1. Exclude specified columns (target, IDs, etc.)
  df_corr <- df %>% dplyr::select(-dplyr::any_of(exclude_cols))
  
  # 2. Identify numeric and numeric-like factor columns
  num_cols <- sapply(df_corr, function(x) {
    is.numeric(x) || (is.factor(x) && all(!is.na(as.numeric(as.character(levels(x))))))
  })
  num_df <- df_corr[, num_cols, drop = FALSE]
  
  # 3. Convert factor columns to numeric codes
  for (col in names(num_df)) {
    if (is.factor(num_df[[col]])) {
      num_df[[col]] <- as.numeric(as.character(num_df[[col]]))
    }
  }
  
  # 4. Remove columns with only one unique non-NA value (constant)
  var_cols <- sapply(num_df, function(x) length(unique(na.omit(x))) > 1)
  num_df   <- num_df[, var_cols, drop = FALSE]
  
  # 5. If at least 2 variables remain, plot the correlation matrix
  if (ncol(num_df) < 2) {
    cat("Not enough numeric columns with variation for correlation plot.\n")
  } else {
    cm <- cor(num_df, use = "pairwise.complete.obs")
    corrplot(
      cm,
      method = "color",
      order  = "hclust",
      tl.cex = 0.8,
      tl.col = "black",
      cl.pos = "r",
      title  = "Correlation Matrix (numeric & factor-coded numeric)",
      mar    = c(0,0,1,0)
    )
  }
}

# Usage example: just call the function with your cleaned dataframe
plot_correlation_matrix(df)

```

```{r}
#Prepare only numeric columns (drop target, IDs, etc.)
exclude_cols <- c(target_variable, "DiagnosisBinary", "ID")
df_num <- df %>%
  dplyr::select(-dplyr::any_of(exclude_cols)) %>%
  dplyr::select(where(is.numeric))

#Compute correlation object
corr_obj <- correlate(df_num)

# 4) Plot full correlation heatmap
plot(corr_obj)
```

```{r}
#make a seperate file for function which we can import and run in the run file
#explore mire dlookr to show efficiency
dlookr_understand <- diagnose(df)
print(dlookr_understand)
```


Plotting distribution of target variable
```{r}
# Plot distribution of target/class variable to check for class imbalance
plot_target_distribution <- function(df, target_col) {
  # Check if the target column exists in the dataframe
  if (!(target_col %in% names(df))) {
    cat("Target column not found.\n")
    return()
  }
  
  # Create a bar plot of the target column using ggplot2
  print(
    ggplot(df, aes_string(x = target_col)) +
      geom_bar(fill = c("green", "red")) +
      labs(
        title = paste("Class Distribution of", target_col),
        x     = target_col,
        y     = "Count"
      ) +
      theme_minimal()
  )
}

# Call the function using your target variable
plot_target_distribution(df, target_col = target_variable)

```


Plotting stacked bar plots of each categorical feature by target class
```{r}
# Plot proportion (stacked, normalized) of each categorical variable by target class
plot_categorical_vs_target <- function(df, target_col) {
  # 1) Identify all factor (categorical) columns, except the target column
  cat_cols <- setdiff(
    names(df)[sapply(df, is.factor)],
    target_col
  )
  
  # 2) Loop over each categorical column and plot distribution by target
  for (col in cat_cols) {
    print(
      ggplot(df, aes_string(x = col, fill = target_col)) +
        geom_bar(position = "fill") +  # Stacks bars and normalizes to proportions
        labs(
          title = paste("Distribution of", col, "by", target_col),
          x     = col,
          y     = "Proportion"
        ) +
        scale_y_continuous(labels = scales::percent) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }
}

# Usage: call with your target variable
plot_categorical_vs_target(df, target_col = target_variable)

```

Chi-squared tests between all factor variables and the target
```{r}
#create a table and cvs export for dissertation support 
# Run chi-squared tests between each factor/categorical column and the target
chi_squared_tests <- function(df, target_col) {
  # 1) Get names of all factor columns
  cat_cols <- names(df)[sapply(df, is.factor)]
  
  # 2) For each categorical variable (except the target)
  for (col in cat_cols) {
    if (col != target_col) {
      tbl <- table(df[[col]], df[[target_col]])
      # Only test if at least 2 rows/columns in table (needed for chisq.test)
      if (nrow(tbl) > 1 & ncol(tbl) > 1) {
        test <- suppressWarnings(chisq.test(tbl))
        cat(
          "\nChi-squared test for:", col, "vs", target_col,
          "\n   p-value:", round(test$p.value, 4),
          ifelse(test$p.value < 0.05, "IS significant\n", "IS NOT significant\n")
        )
      } else {
        cat("\nChi-squared test for:", col, "vs", target_col, 
            "\n   Skipped: Table does not have enough variability.\n")
      }
    }
  }
}

# Run chi-squared tests for all categorical variables against your target
chi_squared_tests(df, target_col = target_variable)
```

```{r}
# One-hot encode all factor columns except the target and any exclusions
one_hot_encode_all_levels <- function(df, target_col, exclude_cols = exclude_variable) {
  # 1) Identify which columns to encode (all factors except target & excluded columns)
  cols_to_encode <- setdiff(
    names(df)[sapply(df, is.factor)],
    c(target_col, exclude_cols)
  )
  
  # 2) Use fastDummies to create dummy variables for all levels (do not remove first dummy)
  df <- fastDummies::dummy_cols(
    df,
    select_columns          = cols_to_encode,
    remove_first_dummy      = FALSE,   # Keep all levels
    remove_selected_columns = TRUE     # Drop original factor columns
  )
  
  return(df)
}

# Apply one-hot encoding to your cleaned data
df_encoded <- one_hot_encode_all_levels(
  df,
  target_col  = target_variable,
  exclude_cols = exclude_variable
)

# Check resulting structure
str(df_encoded)

```
PCA

```{r}
# Run PCA on all numeric columns except the target and excluded columns
run_pca <- function(df, target_col, exclude_cols = exclude_variable) {
  # 1) Determine columns to exclude (target and any additional exclusions)
  cols_to_exclude <- c(target_col, exclude_cols)
  
  # 2) Subset only columns not in the exclude list
  num_df <- df[, !(names(df) %in% cols_to_exclude)]
  
  # 3) Keep only numeric columns
  num_df <- num_df[, sapply(num_df, is.numeric)]
  
  # 4) Scale (standardize) data and run PCA
  pca <- prcomp(num_df, center = TRUE, scale. = TRUE)
  
  # 5) Print explained variance for each principal component
  variance <- summary(pca)$importance[2, ]
  cat("Explained variance by component:\n")
  print(variance)
  
  return(pca)
}

# Example: Run PCA on your encoded dataframe
pca_result <- run_pca(
  df_encoded,
  target_col    = target_variable,
  exclude_cols  = exclude_variable
)

```

```{r}
# Plot a scree plot for a PCA result object
plot_pca_scree <- function(pca) {
  # 1) Calculate the proportion of variance explained by each PC
  var_explained <- pca$sdev^2 / sum(pca$sdev^2)
  
  # 2) Create a data frame for plotting
  scree_data <- data.frame(
    PC = seq_along(var_explained),
    Variance = var_explained
  )
  
  # 3) Plot the scree plot with ggplot2
  ggplot(scree_data, aes(x = PC, y = Variance)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "PCA Scree Plot",
      x     = "Principal Component",
      y     = "Proportion of Variance Explained"
    ) +
    theme_minimal()
}

# Call the function to display the scree plot
plot_pca_scree(pca_result)

```

```{r}
# Run PCA and plot cumulative explained variance
run_pca_analysis <- function(df, target_col = target_variable, exclude_cols = exclude_variable) {
  # 1) Select only numeric columns, then drop the target and any excluded columns
  num_df <- df %>%
    dplyr::select(where(is.numeric)) %>%
    dplyr::select(-dplyr::any_of(c(target_col, exclude_cols)))
  
  # 2) Run PCA with centering and scaling
  pca_result <- prcomp(num_df, center = TRUE, scale. = TRUE)
  
  # 3) Print summary of PCA (shows variance explained, etc.)
  print(summary(pca_result))
  
  # 4) Calculate and store the proportion of explained variance (PEV) for each PC
  pca_result_var <- pca_result$sdev^2
  pca_result_PEV <- pca_result_var / sum(pca_result_var)
  
  # 5) Plot cumulative explained variance (scree-style, but cumulative)
  opar <- par(no.readonly = TRUE)   # Save old graphics parameters
  plot(
    cumsum(pca_result_PEV),
    ylim = c(0, 1),
    xlab = 'Principal Component',
    ylab = 'Cumulative Proportion of Explained Variance',
    type = "b",
    col  = "orange",
    main = "Cumulative Explained Variance by PCs"
  )
  abline(h = 0.8, col = 'red', lty = 'dashed')  # 80% threshold reference
  par(opar)   # Restore old graphics parameters
  
  # 6) Return the PCA object for further analysis (scores, loadings, etc.)
  return(pca_result)
}

# Example usage:
pca_result <- run_pca_analysis(
  df_encoded,
  target_col   = target_variable,
  exclude_cols = exclude_variable
)

```

Working on a stratified sampling here as the dataset might be too large to run PCA on less powerful devices

```{r}
# Stratified sampling: creates a balanced sample by class (target variable)
stratified_sample_by_target <- function(df,
                                        target_col,
                                        sizes,
                                        remove_id = TRUE,
                                        id_col    = "ID") {
  # 1) Ensure the target column is a factor
  if (!is.factor(df[[target_col]])) {
    df[[target_col]] <- as.factor(df[[target_col]])
  }
  
  # 2) Get class labels and number of classes
  classes   <- levels(df[[target_col]])
  n_classes <- length(classes)
  
  # 3) Prepare the size vector for strata sampling
  if (length(sizes) == 1) {
    size_vec <- rep(sizes, n_classes)
  } else if (length(sizes) == n_classes) {
    size_vec <- sizes
  } else {
    stop("`sizes` must be either length 1 or equal to the number of classes (", 
         n_classes, ").")
  }
  
  # 4) Perform stratified sampling without replacement
  stratified_index <- sampling::strata(
    df,
    stratanames = target_col,
    size        = size_vec,
    method      = "srswor"
  )$ID_unit
  
  sampled_df <- df[stratified_index, ]
  
  # 5) Optionally drop the ID column
  if (remove_id && id_col %in% names(sampled_df)) {
    sampled_df <- sampled_df %>% dplyr::select(-all_of(id_col))
  }
  
  return(sampled_df)
}

# Example: Get a balanced sample (e.g., 3,000 per class)
df_ready <- stratified_sample_by_target(
  df         = df_encoded,
  target_col = target_variable,
  sizes      = 3260
)

# Check the resulting class proportions (should be balanced)
prop.table(table(df_ready[[target_variable]]))

```


K-means Clustering

```{r}

#Run PCA on the stratified sample instead of the full encoded set
pca_result_ready <- run_pca(
  df         = df_ready,           #stratified sample
  target_col = target_variable,
  exclude_cols = exclude_variable
)

#Decide how many PCs to keep (90% variance considered)
var_explained <- summary(pca_result_ready)$importance[2, ]
cum_var       <- cumsum(var_explained)
threshold     <- 0.90
n_pcs         <- which(cum_var >= threshold)[1]
cat(
  "Retaining", n_pcs, "PCs (",
  round(cum_var[n_pcs] * 100, 1), "% variance)\n"
)

#Considering those top‐PC scores
pca_scores_ready <- pca_result_ready$x[, 1:n_pcs, drop = FALSE]

#Elbow method on the entire stratified sample
fviz_nbclust(
  pca_scores_ready,
  FUN     = kmeans,
  method  = "wss",
  k.max   = 10
) +
  labs(title = "Elbow Method on Stratified Sample") +
  theme_classic()

#Silhouette analysis on the entire stratified sample
k_values        <- 2:9
kmeans_results  <- list()
sil_widths      <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  
  km <- kmeans(pca_scores_ready, centers = k)
  
  # compute silhouette
  sil <- silhouette(km$cluster, dist(pca_scores_ready))
  sil_widths[i] <- mean(sil[, "sil_width"])
  
  #plot
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ")"),
    border = NA,
    cex.names = 0.6
  )
  
  kmeans_results[[paste0("k", k)]] <- km
}

#Compare average silhouette widths to pick the best k
sil_df <- data.frame(
  k   = k_values,
  ASW = sil_widths
)
print(sil_df)

# Build silhouette summary
sil_df <- data.frame(
  k   = k_values,
  ASW = sil_widths
)
```


```{r}
# 1. Pick the best number of clusters (k) based on silhouette analysis
best_k <- sil_df$k[which.max(sil_df$ASW)]
cat("Best k by average silhouette:", best_k, "\n")

# 2. Re-run k-means clustering using the optimal k on the top principal components
km_best <- kmeans(pca_scores_ready, centers = best_k)

# 3. Prepare a data frame for plotting: use first two PCs and cluster assignments
plot_df <- data.frame(
  PC1     = pca_scores_ready[, 1],
  PC2     = pca_scores_ready[, 2],
  cluster = factor(km_best$cluster)
)

# 4. Scatter plot the clusters colored by their cluster assignment
ggplot(plot_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("K-means Clustering (k = ", best_k, ") on First Two PCs"),
    x     = "Principal Component 1",
    y     = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```
------K-means Clustering without PCA
```{r}
# 1. Prepare data: remove target/excluded columns and apply min-max scaling
orig_clust_df <- df_ready %>%
  dplyr::select(-dplyr::any_of(c(target_variable, exclude_variable)))
orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

# 2. K-means silhouette analysis for k = 2 to 10
sil_widths_km <- numeric()
kmeans_results <- list()
for (k in 2:10) {
  km <- kmeans(orig_clust_minmax, centers = k)
  kmeans_results[[k]] <- km
  sil <- silhouette(km$cluster, dist(orig_clust_minmax))
  sil_widths_km[k] <- mean(sil[, "sil_width"])
  # Plot silhouette for each k
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ") - K-means (Original Features)"),
    border = NA,
    cex.names = 0.6
  )
}

# 3. Silhouette summary table for all k
sil_df_km <- data.frame(
  k   = 2:10,
  ASW = sil_widths_km[2:10]
)
print(sil_df_km)

# 4. Plot average silhouette width vs k
plot(sil_df_km$k, sil_df_km$ASW, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Width vs. k (K-means, Original Features)")

# 5. Select best k (highest silhouette width)
best_k_km <- sil_df_km$k[which.max(sil_df_km$ASW)]
cat("Best k by average silhouette width (k-means):", best_k_km, "\n")

# 6. Run final k-means on original data with best k
km_best <- kmeans(orig_clust_minmax, centers = best_k_km)

# 7. Plot clusters on the first two min-max scaled features (no PCA)
km_plot_df <- data.frame(
  Feature1 = orig_clust_minmax[, 1],
  Feature2 = orig_clust_minmax[, 2],
  cluster  = factor(km_best$cluster)
)

ggplot(km_plot_df, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("K-means Clustering (k = ", best_k_km, ") on First Two Features"),
    x     = "Feature 1",
    y     = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```


----- Clustering Hierarchical

```{r}
# Hierarchical clustering on the stratified PCA scores

# 1. Compute the Euclidean distance matrix on the PCA scores
hc_dist_full <- dist(pca_scores_ready, method = "euclidean")

# 2. Perform hierarchical clustering using Ward's method
hc_ward_full <- hclust(hc_dist_full, method = "ward.D2")

# 3. Plot the dendrogram (labels omitted for clarity)
plot(
  as.dendrogram(hc_ward_full),
  main   = "Hierarchical Clustering Dendrogram (Ward) on Full Sample",
  ylab   = "Height",
  labels = FALSE,
  hang   = -1
)

# 4. Highlight the best_k clusters on the dendrogram
rect.hclust(hc_ward_full, k = best_k, border = "steelblue")

# 5. Cut the tree into best_k clusters
hc_clusters_full <- cutree(hc_ward_full, k = best_k)

# 6. Prepare data for scatter plot: first two principal components, colored by cluster
hc_plot_df_full <- data.frame(
  PC1     = pca_scores_ready[, 1],
  PC2     = pca_scores_ready[, 2],
  cluster = factor(hc_clusters_full)
)

# 7. Scatter plot of clusters on first two PCs
ggplot(hc_plot_df_full, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k = ", best_k, ") on Full Stratified PCs"),
    x     = "Principal Component 1",
    y     = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )

```


----- Clustering Hierarchical without PCA
```{r}
# 1. Prepare data: remove target/excluded columns and apply min-max scaling
orig_clust_df <- df_ready %>%
  dplyr::select(-dplyr::any_of(c(target_variable, exclude_variable)))
orig_clust_minmax <- as.data.frame(lapply(orig_clust_df, function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}))

# 2. Compute distance matrix and hierarchical clustering
hc_dist_orig <- dist(orig_clust_minmax, method = "euclidean")
hc_ward_orig <- hclust(hc_dist_orig, method = "ward.D2")

# 3. Silhouette analysis for k = 2 to 10
sil_widths_hc <- numeric()
for (k in 2:10) {
  clusters <- cutree(hc_ward_orig, k)
  sil <- silhouette(clusters, hc_dist_orig)
  sil_widths_hc[k] <- mean(sil[, "sil_width"])
  # Plot silhouette for each k
  plot(
    sil,
    main = paste("Silhouette Plot (k =", k, ") - Hierarchical (Original Features)"),
    border = NA,
    cex.names = 0.6
  )
}

# 4. Silhouette summary table for all k
sil_df_hc <- data.frame(
  k   = 2:10,
  ASW = sil_widths_hc[2:10]
)
print(sil_df_hc)

# 5. Plot average silhouette width vs k
plot(sil_df_hc$k, sil_df_hc$ASW, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Width vs. k (Hierarchical, Original Features)")

# 6. Select best k (highest silhouette width)
best_k_hc <- sil_df_hc$k[which.max(sil_df_hc$ASW)]
cat("Best k by average silhouette width (hierarchical):", best_k_hc, "\n")

# 7. Dendrogram with rectangles for chosen clusters
plot(
  as.dendrogram(hc_ward_orig),
  main   = paste0("Hierarchical Clustering Dendrogram (k = ", best_k_hc, ")"),
  ylab   = "Height",
  labels = FALSE,
  hang   = -1
)
rect.hclust(hc_ward_orig, k = best_k_hc, border = "steelblue")

# 8. Assign clusters for the best k
hc_clusters_orig <- cutree(hc_ward_orig, k = best_k_hc)

# 9. Plot clusters on the first two min-max scaled features (no PCA)
hc_plot_df_orig <- data.frame(
  Feature1 = orig_clust_minmax[, 1],
  Feature2 = orig_clust_minmax[, 2],
  cluster  = factor(hc_clusters_orig)
)

ggplot(hc_plot_df_orig, aes(x = Feature1, y = Feature2, color = cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    title = paste0("Hierarchical Clustering (k = ", best_k_hc, ") on First Two Features"),
    x     = "Feature 1",
    y     = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title      = element_text(hjust = 0.5)
  )
```

```{r}
library(dplyr)

# A general helper: 
#   - df:       your data.frame or tibble  
#   - target:   the name of the column you care about (as a string)  
print_target_counts <- function(df, target) {
  df %>% 
    # count by the target column, even if it's a string
    count(!!sym(target)) %>%  
    # rename the count column to something meaningful
    rename(n = n, !!sym(target) := !!sym(target)) %>%  
    # print to console
    print()
}

# Example usage:
# If your cleaned data is in df1_clean:
print_target_counts(df, "Alzheimer.s.Diagnosis")

# If it’s in a merged dataset called `all_df`:
# print_target_counts(all_df, "Alzheimer.s.Diagnosis")

```

